---
title: "Lab 1"
author: "Sam Burch"
date: "2023-02-04"
output:
  pdf_document: default
  html_document: default
---

***Question 1***
``` {r}
library(Lahman)
library(broom)
library(tidyverse)

Teams = as_tibble(Teams)

df_1 = Teams |> 
  select(yearID, W, L, G, AB, H, X2B, X3B, HR, BB, HBP, SF, HA, HRA,
         BBA, SOA, IPouts, FP, R, RA) |> 
  filter(yearID >= 1900) |> 
  replace_na(list(HBP =0, SF = 0)) |> 
  mutate(RD = (R - RA) / G,
         X1B = H - (X2B + X3B + HR)) |> 
  mutate(OBP = (H + BB + HBP) / (AB + BB + HBP + SF)) |> 
  mutate(SLG = (X1B + 2*X2B + 3*X3B + 4*HR) / AB) |> 
  mutate(OPS = OBP + SLG) |> 
  mutate(IP = IPouts / 3) |> 
  mutate(WHIP = (BBA + HA) / IP) |> 
  mutate(FIP = (13*HRA + 3*BBA - 2*SOA) / IP)
df_1
  
mlr_1 = lm(R ~ OPS + WHIP + FP, data = df_1)
summary(mlr_1)
```
R^2 here is ok, but let us now test for diagnostics. <br>

**Diagnostics**
``` {r}
pairs(df_1 |> dplyr::select(R, OPS, WHIP, FP))
```
The linearity assumption holds, as there is clearly a linear relationship between runs and each predictor in the model.

``` {r}
par(mfrow = c(2, 2))
plot(mlr_1)
```
As we check that the errors are normally distributed with constant variance and mean 0, only mean 0 seems to suffice. There seems to be deviations from normality and constant variance assumptions. Also, a large right tail in residuals.

**Problematic Residuals**
``` {r}
par(mfrow = c(1, 1))
qqnorm(residuals(mlr_1))
qqline(residuals(mlr_1))
abline(h = -50, lty = 2, col = "red")
```
It seems like certain residuals are causing the normality and constant variance assumptions to fail.

``` {r}
df_1 = augment(mlr_1, df_1)

df_1 |> 
  filter(.resid < -100) |> 
  group_by(yearID) |> 
  summarise(n())

df_1 |> 
  filter(.resid < -50) |> 
  group_by(yearID) |> 
  summarise(n = n()) |> 
  filter(n >= 10)
```
These years are where the issues are coming from. Let us try to fix this problem by adding in categorical variables for these years.

**Fixed Model**
``` {r}
df_1 = df_1 |> 
  mutate(year_1918 = ifelse(yearID == 1918, 1, 0),
         year_1919 = ifelse(yearID == 1919, 1, 0),
         year_1981 = ifelse(yearID == 1981, 1, 0),
         year_1994 = ifelse(yearID == 1994, 1, 0),
         year_1995 = ifelse(yearID == 1995, 1, 0),
         year_2020 = ifelse(yearID == 2020, 1, 0))


mlr_2 = lm(R ~ OPS + WHIP + FP + year_1918 + year_1919 + year_1981 + year_1994 + year_1995 + year_2020, data = df_1)
summary(mlr_2)
```
Here, the R^2 is much higher, which is a good sign!

``` {r}
par(mfrow = c(2, 2))
plot(mlr_2)
```
All assumptions met now. Thus, this is a much better model than before.

What went wrong is these outlier seasons, due to different circumstances, had drastically different predictor values related to runs. So, after adding in the categorical variables for such years, this problem was solved!


**Principled Rescaling / New Model**
``` {r}
teams_list = Teams |> 
  select(yearID, W, L, G, AB, H, X2B, X3B, HR, BB, HBP, SF, HA, HRA,
         BBA, SOA, IPouts, FP, R, RA) |> 
  replace_na(list(HBP = 0, SF = 0)) |> 
  mutate(RD = (R - RA) / G) |> 
  mutate(X1B = H - (X2B + X3B + HR)) |> 
  mutate(OBP = (H + BB + HBP) / (AB + BB + HBP + SF)) |> 
  mutate(SLG = (X1B + 2*X2B + 3*X3B + 4*HR) / AB) |> 
  mutate(OPS = OBP + SLG) |> 
  mutate(IP = IPouts / 3) |> 
  mutate(WHIP = (BBA + HA) / IP) |> 
  split(Teams$yearID) |> 
  lapply(avgOBP = (sum(H) + sum(BB) + sum(HBP)) / (sum(AB) + sum(BB) + sum(HBP) + sum(SF)),
         avgSLG = (sum(X1B) + 2*sum(X2B) + 3*sum(X3B) + 4*sum(HR)) / sum(AB),
         avgOPS = avgOBP + avgSLG,
         avgWHIP = (sum(BBA) + sum(HA)) / sum(IP),
         #avgFP here is only rough approximation here
         avgFP = mean(FP), 
         mutate) |>
  lapply(OPSscale = OPS / avgOPS,
         WHIPscale = avgWHIP / WHIP,
         FPscale = avgFP / FP,
         mutate)

df_2 = do.call('rbind', teams_list) 

df_2 = df_2 |> 
  filter(yearID >= 1900)


mlr_3 = lm(RD ~ OPSscale + WHIPscale + FPscale, data = df_2)
summary(mlr_3)


par(mfrow = c(2, 2))
plot(mlr_3)
```
The rescaling allowed for a better model compared to the one in the notes (RD ~ OPS + WHIP + FP) because the R^2 is higher and all the model assumptions are met. This rescaling allowed the predictors to be accounted for their specific years. Meaning, if a year is different from the overall average, this accounts for that difference.




***Question 2***

**Batting**

``` {r}
Batting = as_tibble(Batting)
People = as_tibble(People)

batters_3 = People |> 
  filter(playerID == 'troutmi01' |
         playerID == 'goldspa01' |
         playerID == 'justida01') |> 
  dplyr::select(playerID, nameFirst, nameLast) |> 
  pull(playerID)

bat_seasons = Batting |> 
  filter(playerID %in% batters_3) |> 
  group_by(playerID, yearID) |>
  mutate(X1B = H - (X2B + X3B + HR)) |> 
  mutate(OBP = (H + BB + HBP) / (AB + BB + HBP + SF)) |> 
  mutate(SLG = (X1B + 2*X2B + 3*X3B + 4*HR) / AB) |> 
  mutate(OPS = OBP + SLG) |> 
  mutate(SB_pct = (SB / (SB + CS))) |> 
  mutate(season = case_when((playerID == 'justida01' & yearID == 1989) ~ 1,
                            (playerID == 'justida01' & yearID != 1989) ~ (yearID - 1988),
                            (playerID == 'goldspa01' & yearID == 2011) ~ 1,
                            (playerID == 'goldspa01' & yearID != 2011) ~ (yearID - 2010),
                            (playerID == 'troutmi01' & yearID == 2011) ~ 1,
                            (playerID == 'troutmi01' & yearID != 2011) ~ (yearID - 2010),
  )) |>
  dplyr::select(playerID, yearID, season, G:SO, SB_pct, OBP, SLG, OPS)
bat_seasons



Batting |> 
  filter(playerID %in% batters_3) |> 
  mutate(X1B = H - (X2B + X3B + HR)) |> 
  group_by(playerID) |>
  summarise(seasons = n(),
            G = sum(G),
            AB = sum(AB),
            R = sum(R),
            H = sum(H),
            X2B = sum(X2B),
            X3B = sum(X3B),
            HR = sum(HR),
            RBI = sum(RBI),
            SB = sum(SB),
            CS = sum(CS),
            BB = sum(BB),
            SO = sum(SO),
            
            X1B = sum(X1B),
            HBP = sum(HBP),
            SF = sum(SF)
            ) |> 
  mutate(OBP = (H + BB + HBP) / (AB + BB + HBP + SF)) |> 
  mutate(SLG = (X1B + 2*X2B + 3*X3B + 4*HR) / AB) |> 
  mutate(OPS = OBP + SLG) |> 
  mutate(SB_pct = (SB / (SB + CS))) |> 
  dplyr::select(playerID, seasons, G:SO, OBP:SB_pct)


ggplot(bat_seasons, aes(x = season, y = OPS)) +
  labs(
    x = 'Seasons',
    y = 'OPS',
    title = "OPS Career Trajectories",
    subtitle = 'Paul Goldschmidt, David Justice, and Mike Trout',
    caption = 'By: Sam Burch  |  Data @Lahman'
  ) +
  geom_point(aes(size = AB, color = playerID, alpha = .8)) +
  geom_line(aes(color = playerID, alpha = .8), data = bat_seasons |> filter(playerID == 'justida01')) +
  geom_line(aes(color = playerID, alpha = .8), data = bat_seasons |> filter(playerID == 'goldspa01')) +
  geom_line(aes(color = playerID, alpha = .8), data = bat_seasons |> filter(playerID == 'troutmi01')) +
  geom_hline(yintercept =  .729, color = 'darkgrey', linetype = 2)
```
Note: The dotted grey line represents the approximate (modern) average OPS in MLB -- .729.

We can see just how great all three batters have been! Basically every year these batters have been in the league, they've had a great OPS -- well above average. Trout's even been hovering around 1.1 OPS the last few years, which is absurd. For Justice, this graph shows the fall off (as he got older). Lastly, it only took till their second year for all three to be very good hitters.


**Pitching**

``` {r}
Pitching = as_tibble(Pitching)


pitchers_3 = People |> 
  filter(playerID == 'bumgama01' |
           playerID == 'kershcl01' |
           playerID == 'riverma01') |> 
  dplyr::select(playerID, nameFirst, nameLast) |>
  pull(playerID)


pitch_seasons = Pitching |> 
  filter(playerID %in% pitchers_3) |> 
  group_by(playerID, yearID) |>
  mutate(IP = IPouts / 3) |> 
  mutate(WHIP = (BB + H) / IP) |> 
  mutate(SO_per_9 = SO / (IP * 9)) |> 
  mutate(SO_per_BB = SO / BB) |> 
  mutate(season = case_when((playerID == 'riverma01' & yearID == 1995) ~ 1,
                            (playerID == 'riverma01' & yearID != 1995) ~ (yearID - 1994),
                            (playerID == 'kershcl01' & yearID == 2008) ~ 1,
                            (playerID == 'kershcl01' & yearID != 2008) ~ (yearID - 2007),
                            (playerID == 'bumgama01' & yearID == 2009) ~ 1,
                            (playerID == 'bumgama01' & yearID != 2009) ~ (yearID - 2008),
                            )) |> 
  dplyr::select(playerID:yearID, season, W:L, IPouts:SO, HBP, ERA, WHIP, 
                SO_per_9, SO_per_BB, IP)
pitch_seasons


Pitching |> 
  filter(playerID %in% pitchers_3) |> 
  mutate(IP = IPouts / 3) |> 
  group_by(playerID) |> 
  summarise(
    seasons = n(),
    W = sum(W),
    L = sum(L),
    IPouts = sum(IPouts),
    H = sum(H),
    ER = sum(ER),
    HR = sum(HR),
    BB = sum(BB),
    HBP = sum(HBP),
    SO = sum(SO),
    
    IP = sum(IP)
  ) |> 
  mutate(ERA = (9 * ER) / IP) |> 
  mutate(WHIP = (BB + H) / IP) |> 
  mutate(SO_per_9 = SO / (IP * 9)) |> 
  mutate(SO_per_BB = SO / BB) |> 
  dplyr::select(-IP)


People |> 
  filter(playerID == 'bumgama01' |
           playerID == 'kershcl01' |
           playerID == 'riverma01') |> 
  dplyr::select(playerID, nameFirst, nameLast) 



ggplot(pitch_seasons, aes(season, WHIP)) +
  labs(
    x = 'Seasons',
    y = 'WHIP',
    title = "WHIP Career Trajectories",
    subtitle = 'Madison Bumgarner, Clayton Kershaw, and Mariano Rivera',
    caption = 'By: Sam Burch  |  Data @Lahman'
  ) +
  geom_point(aes(size = IP, color = playerID, alpha = .8)) +
  geom_line(aes(color = playerID, alpha = .8), data = pitch_seasons |> filter(playerID == 'bumgama01')) +
  geom_line(aes(color = playerID, alpha = .8), data = pitch_seasons |> filter(playerID == 'riverma01')) +
  geom_line(aes(color = playerID, alpha = .8), data = pitch_seasons |> filter(playerID == 'kershcl01')) +
  geom_hline(yintercept =  1.30, color = 'darkgrey', linetype = 2)
```
Note: The dotted grey line represents the approximate (modern) average WHIP in MLB -- 1.3. Also, the lower the WHIP, the better.

Again, the brilliant play of these guys is clearly seen. Specifically, Kershaw and Rivera have shown elite peaks -- with seasons dipping below .8 WHIP. For Bumgarner, his play (while still good) has been more up and down. On top of that, it is interesting how it took a few seasons for these pitchers to become really good. Maybe that suggests pitchers take longer to develop than hitters; however, these graphs don't necessarily prove that.



***Question 3***
Bob Gibson's 1968 season.

**a)**
``` {r}
Pitching |> 
  filter(playerID == 'gibsobo01', yearID == '1968') |> 
  select(GS, CG)
28/34
```
80% of the 34 games Gibson started (this season) were completed by him.


**b)**
``` {r}
Pitching |> 
  filter(playerID == 'gibsobo01', yearID == '1968') |> 
  select(BB, SO)
268/62
```
About 4.3 strikeouts per walk.


**c)**
``` {r}
Pitching |> 
  filter(playerID == 'gibsobo01', yearID == '1968') |> 
  mutate(IP = IPouts / 3) |> 
  select(IP)
```
Just over 300 IP.


**d)**
``` {r}
Pitching |> 
  filter(playerID == 'gibsobo01', yearID == '1968') |> 
  mutate(IP = IPouts / 3) |> 
  mutate(WHIP = (BB + H) / IP) |> 
  select(WHIP)
```
About .85 WHIP.


***Question 4***
Jim Bunning's perfect game on Father's Day.

``` {r}
library(retrosheet)
retro_1964 = as_tibble(getRetrosheet(type = 'game', year = 1964))
```

**a)**
``` {r}
retro_1964 |> 
  filter(Date == '19640621',
         VisTm == 'PHI',
         HmTm == 'NYN',
         DblHdr == '1') |> 
  select(Duration)
```
The game was 2 hours and 19 minutes long.


**b)**
``` {r}
retro_1964 |> 
  filter(Date == '19640621',
         VisTm == 'PHI',
         HmTm == 'NYN') |> 
  select(Attendance)
```
The attendance value is equal to 0 because it is a double header. The attendance for the day is recorded for the second game that day, 32026.

**c)**
``` {r}
retro_1964 |> 
  filter(Date == '19640621',
         VisTm == 'PHI',
         HmTm == 'NYN',
         DblHdr == '1') |> 
  select(VisD, VisT, VisHR) |> 
  sum()
```
Three extra base hits by the Phillies.


**d)**
``` {r}
retro_1964 |> 
  filter(Date == '19640621',
         VisTm == 'PHI',
         HmTm == 'NYN',
         DblHdr == '1') |> 
  mutate(Vis_OBP = (VisH + VisBB + VisHBP) / (VisAB + VisBB + VisHBP + VisSF)) |> 
  select(Vis_OBP)
```
.333 OBP by the Phillies.


